Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2016-02-04 paper.tex
address reviewer 1 comments

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps the most complete published comparison performed.  Despite this,
various results from the simulators were often significantly different due to
inconsistency in modeling decisions involving reprocessing strategies,
refueling behavior, reactor end-of-life handling, etc.  This work identifies
and quantifies the effects of selected modeling choices that may sometimes be
taken for granted in the fuel cycle modeling domain.  Four scenarios are
compared using combinations of either fleet-based or individually modeled reactors with
either monthly or quarterly (3-month) time steps.  The scenarios approximate a
transition from the current U.S. once-through light water reactor (LWR) fleet to a
full sodium fast reactor (SFR) fuel cycle. The Cyclus fuel cycle simulator's
plug-in facility capability along with its market-like dynamic material
routing allow it to be used as a level playing field for comparing the
scenarios.  When under supply-constraint pressure, the four cases exhibit
noticeably different behavior.  Fleet-based modeling is more efficient in
supply-constrained environments at the expense of losing insight on issues
such as realistically suboptimal fuel distribution and challenges in reactor
refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments.  Large simulations
with fleet-based reactors run much more quickly than their individual reactor
counterparts.  Gaining a better understanding of how these and other modeling
choices affect fuel cycle dynamics will enable making more deliberate
decisions with respect to trade-offs such as computational investment vs.
realism.
}

@@

@@

@@

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
likely be due to quantized shutdown.

@@

Figure \ref{fig:power} shows an overall view of the generated power over time
for each of the four cases.  Figure \ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve magnifying some
interesting differences between the four cases.  For the first 100 years, the
four cases behave somewhat similarly, although the individually modeled
reactors in cases MI and QI actually go offline and back online for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in case QF is slightly
lower than in case MF during the fuel shortage years. This is primarily due to
the \emph{quantized shutdown effect}.  The longer 3-month time step in case QF 
results in reactor capacity going offline longer than necessary.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepancy
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the misalignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

Increasing time step duration increases facilities' standing inventory
requirement during supply constrained periods.  Facilities can provide no more
inventory than they can hold on a single time step.  So increasing time step
duration for a facility with unconstrained supply generally means higher
standing inventories are required in order to achieve equal throughput.
Facilities with large inventory buffer that are unconstrained are not affected
significantly by the time step duration when operating in isolation.  However,
even when facilities do not have individual inventory or throughput limits,
the time step can still have a significant effect. The \emph{inventory drawdown
effect} can impact the aggregate throughput of supply-constrained recycle loops by affecting
minimum bounds on idling material quantities.

The impact of modeling assumptions depends heavily on the metrics of interest.
For example, someone interested in fuel-shortage driven reactor outages would
observe a factor of four span among the four cases shown in the detail
comparison.  However, someone more interested in the overall transition
mechanics would not see as significant a discrepancy.  In an unconstrained
supply setting, varying the time step duration or discretization of facility
modeling have a relatively minor impact on the overall simulation results.
Differences become more significant in material constrained regimes.
Robustness in constrained regimes is especially useful in the context of
sensitivity and optimization analysis.

The value of realism in fuel cycle modeling is greatly be enhanced by an
associated understanding of how outcomes are affected by added fidelity.
Cyclus' flexibility for accommodating different modeling choices uniquely
enables many interesting comparisons.  Those in the field of fuel cycle
analysis in general should be cognizant of how modeling choices are affecting
results and conclusions through exercises like the one here.  Even if the
outcome is that a particular modeling decision has little impact on results,
work in this direction has great potential to improve the integrity of fuel
cycle analysis generally.

>>>>

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps the most complete published comparison performed.  Despite this,
various results from the simulators were often significantly different due to
inconsistencies in modeling decisions involving reprocessing strategies,
refueling behavior, reactor end-of-life handling, etc.  This work identifies
and quantifies the effects of selected modeling choices that may sometimes be
taken for granted in the fuel cycle simulation domain.  Four scenarios are
compared using combinations of either fleet-based or individually modeled
reactors with either monthly or quarterly (3-month) time steps.  The scenarios
approximate a transition from the current U.S. once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle. The Cyclus
fuel cycle simulator's plug-in facility capability along with its market-like
dynamic material routing allow it to be used as a level playing field for
comparing the scenarios.  When under supply-constraint pressure, the four
cases exhibit noticeably different behavior.  Fleet-based modeling is more
efficient in supply-constrained environments at the expense of losing insight
on issues such as realistically suboptimal fuel distribution and challenges in
reactor refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments resulting in much lower
standing inventories of separated Pu.  Large simulations with fleet-based
reactors run much more quickly than their individual reactor counterparts.
Gaining a better understanding of how these and other modeling choices affect
fuel cycle dynamics will enable making more deliberate decisions with respect
to trade-offs such as computational investment vs.  realism.
}

@@

This work identifies and quantifies the effects of selected modeling choices
that may sometimes be taken for granted in the fuel cycle simulation domain.
Four scenarios are compared using combinations of either fleet-based or
individually modeled reactors with either monthly or quarterly (3-month) time
steps.  Time step duration and the degree facility discretization are design
decisions that are handled in a variety of unique ways among nuclear fuel
cycle simulators, and a better understanding of their effects can serve to
increase confidence in results and can help make inter-simulator comparisons
more meaningful.  Reducing error and noise artifacts associated with modeling
decisions will provide a more stable foundation for understanding the
significance of external market, political, and social forces that often
dominate the industry in the real world.

@@

Informal comparison of results from the VISION and Cyclus simulators when
running equivalent scenarios revealed some surprising discrepencies.  One of
the most apparent was a large difference in separated Pu levels between the
two simulators --- something that some in the DOE and nuclear community are
very sensitive to.  For one particular scenario, the Cyclus results also
showed no fuel shortages for reactors, while the VISION results indicated
noticable shortages.  This observation prompted a closer investigation to
determine if the differences were caused by mistakes in simulator input, bugs
in the simulator(s), or differing modeling choices between the simulators.
This investigation eventaully developed into this work which shows important
effects of some modeling choices that are operative in many fuel cycle
simulators.

@@

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
partly be due to quantized shutdown.

@@

Figure \ref{fig:power} shows an overall view of the generated power over time
for each of the four cases.  Figure \ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve magnifying some
interesting differences between the four cases.  For the first 100 years, the
four cases behave somewhat similarly, although the individually modeled
reactors in cases MI and QI actually go offline and back online for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in case QF is slightly
lower than in case MF during the fuel shortage years. This is partly due to
the \emph{quantized shutdown effect}.  The longer 3-month time step in case QF 
results in reactor capacity going offline longer than necessary.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} causes part of the discrepancy
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the misalignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

Increasing time step duration increases facilities' standing inventory
requirement during supply constrained periods.  Here we observe a factor of
three difference between standing inventories of separated Pu for a large
portion of the simulation --- something certain analyses are very sensitive
to.  Facilities can provide no more inventory than they can hold on a single
time step.  So increasing time step duration for a facility with unconstrained
supply generally means higher standing inventories are required in order to
achieve equal throughput.  Facilities with large inventory buffer that are
unconstrained are not affected significantly by the time step duration when
operating in isolation.  However, even when facilities do not have individual
inventory or throughput limits, the time step can still have a significant
effect. The \emph{inventory drawdown effect} can impact the aggregate
throughput of supply-constrained recycle loops by affecting minimum bounds on
idling material quantities.

The impact of modeling assumptions depends heavily on the metrics of interest.
For example, someone interested in fuel-shortage driven reactor outages would
observe a factor of four span among the four cases shown in this comparison
exercise.  However, someone more interested in the overall transition
mechanics would not see as significant a discrepancy.  In an unconstrained
supply setting, varying the time step duration or discretization of facility
modeling has a smaller impact on the overall simulation results.
Differences become more significant in material constrained regimes.
Robustness in constrained regimes can be especially useful in the context of
sensitivity and optimization analysis.

The value of realism in fuel cycle modeling is greatly enhanced by an
associated understanding of how outcomes are affected by added fidelity.
Cyclus' flexibility for accommodating different modeling choices uniquely
enables many interesting comparisons.  Those in the field of fuel cycle
analysis in general should be cognizant of how modeling choices are affecting
results and conclusions through exercises like the one here.  Even if the
outcome is that a particular modeling decision has little impact on results,
efforts to measure and document these impacts can provide confidence in the
appropriateness of modeling choices, reducing the need to rely on intuition
for cycle analysis work.
